NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-11-05 Manuscript/manuscriptPGPE.tex
Update manuscript

Here, we use open collaboration and order parameters of glycerol backbone and headgroup
to evaluate the accuracy of PE and PG heagroup structures, and the cation binding affinity
to anionic membranes containing PG lipids in the current MD simulation force fields.
The force field giving the best description for glycerol backbone and headgroup structures
of PC, PS, PG and PE headgroups (CHARMM36) reproduces the essential differences in order parameters
between these headgroups, and therefore enables the analysis of structural differences between the
headgroups.

@@

Simulations also qualitatively reproduce the experimentally observed decrease of POPG
%Experimental data for the
$\beta$-carbon order parameter
%of POPG shows a rapid decrease with increasing
upon addition of CaCl$_2$,
%concentrations up to 10 mM
suggesting that the PG headgroup response is qualitatively captured by simulations, in contrast
to PS in our previous work \cite{antila19}.
and more modest decrease with larger concentrations (Fig. \ref{changesWITHCaClPG}) \cite{borle85}.
This behaviour is similar to that of $\beta$-carbon order parameters of POPC,
but essentially different than observed for POPS, where $\beta$-carbon order parameters increases with
addition of calcium \cite{antila19}. Experimentally measured changes of PG $\alpha$-carbon order parameters upon addition of
calcium are not available.
%upon addition of calcium in experiments,
%exhibiting 
%The behaviour is qualitatively different than in the $\beta$-carbon of PS lipids which increase upon addition of CaCl$_2$.
Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
even thought the binding affinity was too large based on the comparison of PC headgroup order parameter changes with experiments.
While applying ECC to Lipid17 improved the PC headgroup order parameter response and binding affinity,
the response of PG $\beta$-carbon order parameter to calcium is too weak in this model.
The response of PG $\alpha$-carbon order parameters to CaCl$_2$ differs between force fields,
but experimental data to evaluate these predictions is not available. 
\todo{To be finished once we have the new CHARMM simulations and conformational changes of PG analyzed.}
% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.

@@

\todo{We still need more data to finish the discussion. More detailed discussion is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}

>>>>

Here, we use open collaboration and order parameters of glycerol backbone and headgroup
to evaluate the accuracy of PE and PG heagroup structures, and the cation binding affinity

to anionic membranes containing PG lipids in the current MD simulation force fields.
The force field giving the best description for glycerol backbone and headgroup structures
of PC, PS, PG and PE headgroups (CHARMM36) reproduces the essential differences in order parameters
between these headgroups, and therefore enables the analysis of structural differences between the
headgroups.

@@

Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
even thought the binding affinity was too large based on the comparison of PC headgroup order parameter changes with experiments.
%While applying ECC to Lipid17 improved the PC headgroup order parameter response and binding affinity,
The response of PG $\beta$-carbon order parameter to calcium is too weak in CHARMM36 and ECC-lipid17 force fields,
but decrease of the order parameter is associated with decreased P-N vector angle in all
force fields. Even thought the change in headgroup orientation upon addition of calcium is smaller
in PG lipids than in PC, the conformational changes in PG are more significant because also
dihedral angle distribution around O$_\alpha$-C$_\alpha$ bond is affected by calcium binding (Fig. ??).
However, because none of the force fields simulatneously capture both the PG conformational ensemble and calcium binding,
more accurate force fields are required for more detailed studies of the effect of ions to PG conformational enembles.
%and PG
%The response of PG $\alpha$-carbon order parameters to CaCl$_2$ differs between force fields,
%but experimental data to evaluate these predictions is not available. 
%Simulations also qualitatively reproduce the experimentally observed decrease of POPG
%%Experimental data for the
%$\beta$-carbon order parameter
%%of POPG shows a rapid decrease with increasing
%upon addition of CaCl$_2$,
%%concentrations up to 10 mM
%suggesting that the PG headgroup response is qualitatively captured by simulations, in contrast
%to PS in our previous work \cite{antila19}.
%and more modest decrease with larger concentrations (Fig. \ref{changesWITHCaClPG}) \cite{borle85}.
%This behaviour is similar to that of $\beta$-carbon order parameters of POPC,
%but essentially different than observed for POPS, where $\beta$-carbon order parameters increases with
%addition of calcium \cite{antila19}. Experimentally measured changes of PG $\alpha$-carbon order parameters upon addition of
%calcium are not available.
%%upon addition of calcium in experiments,
%%exhibiting 
%%The behaviour is qualitatively different than in the $\beta$-carbon of PS lipids which increase upon addition of CaCl$_2$.

% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.

@@

%\todo{We still need more data to finish the discussion. More detailed discussion is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}
